Advances in Artificial Intelligence for Data Visualization: Automate Reading of Diagnostic Plots with Compute Vision Models

### 2023-04-11 Milestone_2/report/report.Rmd
compiled

Modern computer vision models offer a promising solution to this challenge. As a subfield of AI, computer vision with its modern deep learning architectures has successfully resolved numerous critical problems in automation. The development of the convolutional neural network (CNN) by @fukushima_neocognitron_1982 was inspired by the vision processing in living organisms. This architecture was first applied to hand-written number recognition trained with back-propagation by @lecun_backpropagation_1989, making it one of the earliest attempts at successfully extracting information from digital images via self-learning algorithms. The modern computer vision model typically utilizes deep neural networks with convolutional layers [@fukushima_neocognitron_1982], which leverage the hierarchical pattern in data and provide regularized versions of fully-connected layers. This approach downscales and transforms images by summarizing information in a small space. Numerous studies have shown that it can effectively tackle vision tasks, such as image recognition [@rawat_deep_2017]. With the development of graphics processing units and the widespread availability of high-performance personal computers, computer vision research become a new hype in the 21st century. Remarkable achievements, such as computer-aided diagnosis [@lee_image_2015], pedestrian detection [@brunetti_computer_2018], and facial recognition [@emami_facial_2012], have significantly impacted our daily life.

@@

The main objective of this research is to construct an automatic visual inference system to enable the conduction of regression diagnostics visual tests on a large scale. The study will concentrate on three specific projects, namely

@@

>>>>

Modern computer vision models offer a promising solution to this challenge. As a subfield of AI, computer vision with its modern deep learning architectures has successfully resolved numerous critical problems in automation. The development of the convolutional neural network (CNN) by @fukushima_neocognitron_1982 was inspired by the vision processing in living organisms. The modern computer vision model typically utilizes deep neural networks with convolutional layers, which leverage the hierarchical pattern in data and provide regularized versions of fully-connected layers. This approach downscales and transforms images by summarizing information in a small space. Numerous studies have shown that it can effectively tackle vision tasks, such as image recognition [@rawat_deep_2017], computer-aided diagnosis [@lee_image_2015], pedestrian detection [@brunetti_computer_2018], and facial recognition [@emami_facial_2012].

@@

The main objective of this research is to construct an automatic visual inference system capable of conducting visual tests on a large scale in the domain of model diagnostics, particularly regression diagnostics. The study will concentrate on three specific projects, namely

@@

The thesis is structured

### 2023-04-11 Milestone_2/report/report.tex
compiled

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Advances in Artificial Intelligence for Data Visualization: Automate Reading of Diagnostic Plots with Compute Vision Models},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{a4paper, centering, text={16cm,25cm}}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{report.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

Practically, visual inference is conducted via the lineup protocol. The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. It comprises \(m\) randomly positioned plots, where on of them represents the data plot, while the remaining \(m - 1\) plots represent the null plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis \(H_0\). Under \(H_0\), the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is \(1/m\). If the data plot is correctly identified, it provides evidence against \(H_0\).

@@

The workload of individuals could be alleviated by automating repetitive tasks and providing standardized results in a controlled environment. The evaluation of visual tests on a large scale is not feasible without the aid of machines and technology.

Modern computer vision models offer a promising solution to this challenge. As a subfield of AI, computer vision with its modern deep learning architectures has successfully resolved numerous critical problems in automation. The development of the convolutional neural network (CNN) by \textcite{fukushima_neocognitron_1982} was inspired by the vision processing in living organisms. This architecture was first applied to hand-written number recognition trained with back-propagation by \textcite{lecun_backpropagation_1989}, making it one of the earliest attempts at successfully extracting information from digital images via self-learning algorithms. The modern computer vision model typically utilizes deep neural networks with convolutional layers \autocite{fukushima_neocognitron_1982}, which leverage the hierarchical pattern in data and provide regularized versions of fully-connected layers. This approach downscales and transforms images by summarizing information in a small space. Numerous studies have shown that it can effectively tackle vision tasks, such as image recognition \autocite{rawat_deep_2017}. With the development of graphics processing units and the widespread availability of high-performance personal computers, computer vision research become a new hype in the 21st century. Remarkable achievements, such as computer-aided diagnosis \autocite{lee_image_2015}, pedestrian detection \autocite{brunetti_computer_2018}, and facial recognition \autocite{emami_facial_2012}, have significantly impacted our daily life.

@@

\begin{itemize}
\tightlist
\item
  cut this p down, we don't need that many details
\end{itemize}

In the initial phase of this research, a pilot investigation was undertaken to explore the viability of implementing visual inference in regression diagnostics. The pilot study involved 64 participants and focused on utilizing a lineup consisting of 20 plots, wherein a residual plot was embedded along with 19 null plots, simulated using the residual rotation technique. The study considered two primary forms of residual departures, namely non-linearity and heteroskedasticity, within a multiple linear regression context. To enrich the visual features, different fitted value distributions, including normal, uniform, lognormal, and discrete uniform, were also incorporated.

The outcomes of this pilot study demonstrated that the power of visual tests in regression diagnostics increases proportionately with the number of lineup evaluations. Nevertheless, it was observed that conventional tests, such as the F-test for non-linearity and the Breusch-Pagan test for heteroskedasticity, exhibited significantly greater power than the visual tests. While the pilot study successfully provided a comparative analysis of the power of visual and conventional tests, the impact of various factors on the power of visual tests could not be fully analyzed due to inadequate data.

We acknowledge the limitations of this pilot study, particularly in the design of the simulated model, which made it challenging to explicate the involved factors. Furthermore, the methodology utilized for effect size calculation, p-value calculation, and power estimation was incipient. Hence, a follow-up study is necessary to address these limitations and obtain a more comprehensive understanding of the potential of visual inference in regression diagnostics.

In the subsequent year, a similar study was conducted, utilizing simpler simulated models to elicit more distinct visual patterns. Specifically, the non-linearity model was formulated as

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\), \(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\), \(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(He_{j}(.)\) is the \(j\)th-order probabilist's Hermite polynomials, \(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and \(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of the random vector to be \([-k, k]^n\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = \frac{\boldsymbol{x} - min(\boldsymbol{x})}{max(\boldsymbol{x}) - min(\boldsymbol{x})}(2k - k), \quad \text{for} \quad k > 0. 
\end{equation}

The heteroskedasticity model was formulated as

\begin{align} \label{eq:heter-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}). 
\end{align}

The null regression model used to fit the realizations generated by the above models is formulated as

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0 + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

\noindent where \(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\).

Since \(z = O(x^j)\), for \(j > 1\), \(z\) is a higher order term leaves out by the null regression, which will violate the linearity model assumption. Similarly, for \(b \neq 0\), the variance-covariance matrix of the error term \(\boldsymbol{\varepsilon}\) is correlated with the predictor \(\boldsymbol{x}\), which will lead to the presence of heteroskedasticity.

\begin{itemize}
\tightlist
\item
  draft below
\end{itemize}

The effect size was derived using the Kullback-Leibler divergence method, formulated as Equation 3. To account for the interdependence of lineups presented to different participants, p-value calculation, as suggested by VanderPlas et al.~(2021), was adopted. Furthermore, power estimation was simplified through a logistic regression model that lacked an intercept but included a fixed offset to ensure that the power equated to 5\% when the effect size was zero. A detailed description of the experimental design, effect size derivation, p-value calculation, and power estimation can be found in the attached paper.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  summarise the work in the first year
\end{enumerate}

\begin{itemize}
\tightlist
\item
  an experiment to compare visual test with conventional test
\item
  a prototype model for lineup evaluation
\item
  points out the limitation of the work
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  summarise the work in the second year
\end{enumerate}

\begin{itemize}
\tightlist
\item
  refine and reconduct the experiment
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  main findings to date
\end{enumerate}

\begin{itemize}
\tightlist
\item
  findings from the first year
\item
  findings from the second year
\end{itemize}

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Advances in Artificial Intelligence for Data Visualization: Automate Reading of Diagnostic Plots with Compute Vision Models},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{a4paper, centering, text={16cm,25cm}}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{report.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

Practically, visual inference is conducted via the lineup protocol. The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. It comprises \(m\) randomly positioned plots, where one of them represents the data plot, while the remaining \(m - 1\) plots represent the null plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis \(H_0\). An example lineup is provided in Figure \ref{fig:lineup-example}. To compute the \(p\)-value of the visual test, the lineup will be independently presented to a number of participants, asking them to pick the most different plot. Under \(H_0\), the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is \(1/m\). If a large number of participants correctly identify the data plot, the corresponding \(p\)-value will be small, indicating strong evidence against \(H_0\).

\begin{figure}
\centering
\includegraphics{report_files/figure-latex/lineup-example-1.pdf}
\caption{\label{fig:lineup-example}Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot \(2^2 + 2\), exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the \(p\)-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small \(p\)-value would result from a substantial number selecting plot \(2^2 + 2\).}
\end{figure}

@@

Modern computer vision models offer a promising solution to this challenge. As a subfield of AI, computer vision with its modern deep learning architectures has successfully resolved numerous critical problems in automation. The development of the convolutional neural network (CNN) by \textcite{fukushima_neocognitron_1982} was inspired by the vision processing in living organisms. The modern computer vision model typically utilizes deep neural networks with convolutional layers, which leverage the hierarchical pattern in data and provide regularized versions of fully-connected layers. This approach downscales and transforms images by summarizing information in a small space. Numerous studies have shown that it can effectively tackle vision tasks, such as image recognition \autocite{rawat_deep_2017}, computer-aided diagnosis \autocite{lee_image_2015}, pedestrian detection \autocite{brunetti_computer_2018}, and facial recognition \autocite{emami_facial_2012}.

@@

In order to examine the potential applicability of integrating visual inference techniques into regression diagnostics, a pilot study was conducted in the first year with the participation of 64 individuals, which was followed by a formal study involving 443 participants in the subsequent year. The participants were presented with lineups consisting of 20 plots, where a residual plot was embedded along with 19 null plots, drawn with data simulated using the residual rotation technique. The study considered two primary forms of residual departures in multiple linear regression, namely non-linearity and heteroskedasticity. To enrich the visual features, different fitted value distributions, including normal, uniform, lognormal, and discrete, were also incorporated.

The study revealed that conventional residual-based statistical tests are more sensitive to weak residual departures from model assumptions compared to visual tests evaluated by humans, leading to excessive rejections even when downstream analysis and outcomes would not be significantly impacted by the small departures from a good fit. Specifically, conventional tests conclude that the model fit has issues nearly twice as frequently as humans would, and they frequently reject departures in the form of non-linearity and heteroskedasticity that are not visible to humans. These findings emphasize the crucial role of graphical diagnostics and support the integration of visual inference into regression diagnostics. Moreover, they provide a compelling rationale for the development of an automated visual inference system to evaluate lineups of residual plots. The comprehensive details of the research methodology, along with other intriguing findings can be found in the attached paper.
